Towards a standardized evaluation of imputation methodology

### 2022-07-15 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
Add reviewer comments

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper, we propose a move towards a standardized evaluation of imputation methodology. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of  imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems.
\end{abstract}

@@

@@

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. In such studies a complete dataset is usually generated from a statistical model, another model is used to induce missingness, and a set of evaluation criteria is postulated to evaluate the performance of one or more missing data methods. However, no golden standard has been established to evaluate imputation routines and, as a result, the validity and aim of the simulation setups may differ tremendously from one developer to another. Especially with novel imputation methods being propagated from the fields of machine learning and artificial intelligence, the differences may become more pronounced. Although these promising new methods seem to yield even sharper imputations than now-standard (semi-)parametric imputation methods, the comparison may not be fair due to the simulation setup.

The purpose of this paper is threefold: First, to raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as a reviewer for statistical journals. Second, to provide imputation methodologists with a suggested course of action when simulating missing data problems. This suggested approach should identify common ground but is in no way intended as an absolute solution. This identifies the third purpose of this paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties.

@@

@@

After deciding upon a data-generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

@@

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms.}
\label{table:mech}
\begin{tabular}{lll}
\hline
      & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\ \noalign{\smallskip}
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\ \noalign{\smallskip}
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Omitting further evaluation of the imputation-generating process may yield sub-optimal imputations. The evaluation of simulated results relies on the relation between missing data models, imputation models, and analysis models. When an imputation model is able to capture the essence of the true non-response mechanism relative to the analysis model, the models are said to be congenial \citep{meng94}. The congeniality of all imputation models should be assessed, but methods to assess the suitability of imputation models and diagnose misfit often rely on visual inspection of the imputations \citep[see e.g.][]{abayomi2008diagnostics, bond16}. This makes its assessment an infeasible endeavor in many simulation studies. Moreover, the performance of imputation procedures on distributional properties is often ignored too. Even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny.

@@

If the goal is inference--not prediction--then the uncertainty about estimates needs to be properly quantified. To capture this uncertainty, the standard errors of the estimates should be correctly calculated, which requires multiple imputation. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should be studied. After all, the 95\% confidence interval should contain the `true' value at least 95 out of 100 times \citep[][p. 591]{neym34}.

@@

Estimand(s) or other simulation targets should be defined in the context of the study aim and data-generating mechanism. As a general rule, it would be wise to include one or more multivariate estimands. It is a realistic expectation for imputation methods to preserve both marginal and conditional distributions with respect to the comparative truth. Imputation methods that fail to do so, should not be considered general-purpose methods (e.g., mean imputation).

@@

\begin{table}[htb]
\begin{center}
\caption{Suggested univariate missingness proportions.}
\label{table:prop}
\begin{tabular}{lll}
\hline
      & Reasoning for suggested missingness proportion \\
\hline  
10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
      & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
      & the true data inference. Performance of a missing data method should at least be \\
      & acceptable for most missing data problems. \\ \noalign{\smallskip} % & \\
25\%  & This is a fair amount of missingness and will, depending on the observed data \\
      & information, have a noticeable influence on the completed data inference. When \\
      & compared to the condition with 10\% missingness, the inference obtained under \\
      & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
      & should increase), but estimates should still be properly covered and the statistical \\
      & properties of the missing data method should be sound. In practice, at least to our \\
      & at least to our experience in social sciences and official statistics, 25\% \\
      & univariate missingness can easily be considered as a realistic missingness percentage. \\ \noalign{\smallskip} % & \\
50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
      & severely. Depending on how the missingness mechanism interacts with the simulated \\
      & data, some imputation techniques may yield estimates that are under-covered such that \\
      & the completed data inference should not be deemed valid anymore. If a method yields \\
      & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
      & properties of the imputation methodology are sound. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

(ii) The fit of the imputation model may be verified with the help of a posterior predictive check \citep[][]{nguy17, zhao22}. A straightforward posterior predictive check for imputation methodology is the multiple over-imputation of observed data values. If the statistical properties of the over-imputed values are equivalent to those of the observed data values, one could infer that the imputation model fits the observed part of the incomplete data reasonably well. Then, by extension, one could assume that the imputation model might be able to produce good imputations for the missing part of the data too.

@@

(iv) Finally, the plausibility of the imputed values may be evaluated. Plausible imputations--imputations that could be real values if they had been observed--are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausible imputations may be a desired property. One would prefer an imputation technique to yield both valid inference and plausible imputations. It should be studied if an imputation method is prone to deliver such impractical results, and if so, under what conditions. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible values.

@@

@@

We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values \citep[][\S 2.5]{buur18}. However, if a study is aimed at obtaining predictions or classification and inferential validity is not of interest, the RMSE of the predictions may yield valuable information about the methods' efficiency in terms of both accuracy and precision.

>>>>

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper, we propose a move towards a standardized evaluation of imputation methodology. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of  imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems. [TODO: explain how the standardisation leads to more neutrality, but that this is just a set of elements to consider/report, not a complete 'cookbook'.]
\end{abstract}

@@

[TODO: Narrowing the scope of the manuscript to the evaluation of methodology for statistical inference, not prediction or causal inference (although inferential validity encompasses predictive validity and may thus be generalized?). Emphasizing that the ideal evaluation of imputation methodology depends on the aim of the specific study (i.e. how the evaluated methods are used in practice).]

@@

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. In such studies a complete dataset is usually generated from a statistical model, another model is used to induce missingness, and a set of evaluation criteria is postulated to evaluate the performance of one or more missing data methods. However, no gold standard has been established to evaluate imputation routines and, as a result, the validity and aim of the simulation setups may differ tremendously from one developer to another. Especially with novel imputation methods being propagated from the fields of machine learning and artificial intelligence, the differences may become more pronounced. Although these promising new methods seem to yield even sharper imputations than now-standard (semi-)parametric imputation methods, the comparison may not be fair due to the simulation setup.

The purpose of this paper is threefold: First, to raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as a reviewer for statistical journals. Second, to provide imputation methodologists with a suggested course of action when simulating missing data problems. This suggested approach should identify common ground but is in no way intended as an absolute solution. This identifies the third purpose of this paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties. [TODO: explain how the standardisation leads to more neutrality, but that this is just a set of elements to consider/report, not a complete 'cookbook'.]

@@

[TODO: Split up the problems with the simulation setup/evaluation and the problems with the interpretation/extrapolation from these simulation studies. See e.g. \citet{gree17} on cognitive problem of generelization versus methodological problems.]

@@

After deciding upon a data-generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results. [TODO: add nuance because this doesn't always hold, see Tims review!]

@@

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms. [TODO: add refs \citet{seam13, meal15, dore18} for clarification and alternatives outlined in \citet{moha21, scho21, more18}]}
\label{table:mech}
\begin{tabular}{lll}
\hline
      & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\ \noalign{\smallskip}
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\ \noalign{\smallskip}
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Omitting further evaluation of the imputation-generating process may yield sub-optimal imputations. The evaluation of simulated results relies on the relation between missing data models, imputation models, and analysis models. When an imputation model is able to capture the essence of the true non-response mechanism relative to the analysis model, the models are said to be congenial \citep{meng94}. [TODO: change this to being able to embed the analysis procedure and the imputation model within a Bayesian model; the nonresponse mdel may be misspecified while the imputation procedure is congenial.]The congeniality of all imputation models should be assessed, but methods to assess the suitability of imputation models and diagnose misfit often rely on visual inspection of the imputations \citep[see e.g.][]{abayomi2008diagnostics, bond16}. This makes its assessment an infeasible endeavor in many simulation studies. Moreover, the performance of imputation procedures on distributional properties is often ignored too. Even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny.

@@

If the goal is inference--not prediction--then the uncertainty about estimates needs to be properly quantified. To capture this uncertainty, the standard errors of the estimates should be correctly calculated, which requires multiple imputation. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should be studied. After all, the 95\% confidence interval should contain the `true' value at least 95 out of 100 times \citep[][p. 591]{neym34}. [TODO: add nuance about coverage rate > 95\% irt CIW and efficiency.]

@@

Estimand(s) or other simulation targets should be defined in the context of the study aim and data-generating mechanism. As a general rule, it would be wise to include one or more multivariate estimands. [TODO: make this part about the study aim and refer to \citet{pete14} about clear and unambiguous estimand, and to focus on this estimand.] It is a realistic expectation for imputation methods to preserve both marginal and conditional distributions with respect to the comparative truth. Imputation methods that fail to do so, should not be considered general-purpose methods (e.g., mean imputation).

@@

\begin{table}[htb]
\begin{center}
\caption{Suggested univariate missingness proportions. [TODO: change this to 'more than one missingness proportion, depending on the context' and/or add references for the given proportions (e.g. about  that 30\% is a good range for MI).]}
\label{table:prop}
\begin{tabular}{lll}
\hline
      & Reasoning for suggested missingness proportion \\
\hline  
10\%  & Depending on the size of the data, this percentage can be considered as a lower \\
      & bound of realistic evaluation. Anything less than 10\% may be of little influence on \\
      & the true data inference. Performance of a missing data method should at least be \\
      & acceptable for most missing data problems. \\ \noalign{\smallskip} % & \\
25\%  & This is a fair amount of missingness and will, depending on the observed data \\
      & information, have a noticeable influence on the completed data inference. When \\
      & compared to the condition with 10\% missingness, the inference obtained under \\
      & 25\% missingness should be less certain (i.e. confidence/credibility interval width \\
      & should increase), but estimates should still be properly covered and the statistical \\
      & properties of the missing data method should be sound. In practice, at least to our \\
      & at least to our experience in social sciences and official statistics, 25\% \\
      & univariate missingness can easily be considered as a realistic missingness percentage. \\ \noalign{\smallskip} % & \\
50\%  & Performance under 50\% percent simulated missingness will most likely be impacted \\ 
      & severely. Depending on how the missingness mechanism interacts with the simulated \\
      & data, some imputation techniques may yield estimates that are under-covered such that \\
      & the completed data inference should not be deemed valid anymore. If a method yields \\
      & acceptable inference under 50\% MAR missingness, we can determine that the statistical \\
      & properties of the imputation methodology are sound. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

(ii) The fit of the imputation model may be verified with the help of a posterior predictive check \citep[][]{nguy17, zhao22}. A straightforward posterior predictive check for imputation methodology is the multiple over-imputation of observed data values. If the statistical properties of the over-imputed values are equivalent to those of the observed data values, one could infer that the imputation model fits the observed part of the incomplete data reasonably well. Then, by extension, one could assume that the imputation model might be able to produce good imputations for the missing part of the data too. [TODO: add that this is implemented in amelia.]

@@

(iv) Finally, the plausibility of the imputed values may be evaluated. Plausible imputations--imputations that could be real values if they had been observed--are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausible imputations may be a desired property. One would prefer an imputation technique to yield both valid inference and plausible imputations. It should be studied if an imputation method is prone to deliver such impractical results, and if so, under what conditions. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible values. [TODO: add nuance here because valid inference, not plausible imputed values should be the aim.]

@@

% "Note that Neyman's original description of confidence intervals defined the property of randomisation validity as exactly 100(1_alpha) of intervals containing theta.44-46 Confidence validity is the property that the true percentage is at least 100(1_alpha). This latter definition is less well known than the former, with the result that overcoverage and undercoverage are sometimes regarded as similarly bad.47 Of course, randomisation validity would usually be preferred over confidence validity because it implies shorter intervals, but this is not always the case. There are examples of procedures that return both shorter intervals and higher coverage.45, 46" ADEMP paper

@@

We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values \citep[][\S 2.5]{buur18}. However, if a study is aimed at obtaining predictions or classification and inferential validity is not of interest, the RMSE of the predictions may yield valuable information about the methods' efficiency in terms of both accuracy and precision. [TODO: make this an optional bullet point for prediction only: out of sample mean squared prediction error can be compared between methods.]
